Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-27 manuscript/manuscript.Rmd
:pen:

MSC:
- MSC code 1
- MSC code 2

@@

Ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g. conservation goals, natural resource extraction, economic and social constraints.)
Despite the complexity inherent in natural ecosystems, analytically tractable models have long played an essential role in their understanding and management.
...

<!-- Why fisheries-focused examples -->
Global fisheries management has long been a crucible for both the development of theory and best practices in actual management for many reasons. Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to overfishing [@Worm2006, @Costello2016]. 
Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].

@@

```{r}
#| label: conceptual
#| echo: FALSE
#| message: FALSE
#| fig.align: 'center'
#| fig.cap:  '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$). where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$. An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive. For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2. 
#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics. For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO. 
#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.'
#| out.width: '6.5in' 
#| fig.pos: 'H'

knitr::include_graphics("figures/conceptual.png")
```

@@

# Discussion

Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
This provides very interpretable solutions, such as constant mortality policies.
There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.

We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.

In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.

One intuition for why a constant escapement policy performs so well in such a model comes from the anticorrelation between $X$ and $Y$ noted in the results section.
We have observed in a variety of policies, both classical as well as some obtained by training DRL agents.
In all of these, the pattern has been present.
In fact, ‘‘crashes’’ in the $X$ population are almost invariably shortly preceeded by dramatic increases in $Y$.
We hypothesize that this effectively leads to a dimensionality reduction: once the value of $X$ is known, along a path traced by the control system, then $Y$ may be roughly predicted.
A similar, although weaker, anticorrelation is also observed between $X$ and $Z$.
These two informal observations point at an effective simplification of the prediction process: the values of $Y$ and $Z$ are much less impactful to policy decisions than the value of $X$ itself.
We have not yet pushed further to establish a more precise picture of this dynamical simplification.

This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].

This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.

>>>>

#MSC:
#- MSC code 1
#- MSC code 2

@@

Much effort has been spent grappling with the complexity of our natural world in contrast to relative simplicity of the models we use to understand it.
Now, as we grapple with the manifold consequences of human impacts on that world, we are faced with difficult questions of figuring out how best to respond when our knowledge and models fall so far short of the complexity our actions have disrupted.
Heroic amounts of data and computation are being brought to bear on developing better, more complex, more realistic models of our environments and ecosystems, yet these models remain disconnected or oversimplify the space of actions they seek to inform.
In this paper we explore a different tack. 
Here we explore the potential of a emerging data-driven machine learning technique, not to better capture the complexity of the natural processes, but rather, to embrace the complexity of the potential _policies_ that might be used to manage them.
Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.

<!-- Why fisheries-focused examples -- maybe we can cut/merge this with paragraph above -->
There are myriad examples from climate scenarios to protected areas, we will illustrate this with a specific example in the management of marine fisheries, which is both an important issue to itself and a frequent crucible in which methods of ecological management have methods for ecological management have been fired before extending to other fields, and where the space of possible policies is arguably already more richly developed than most.
Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016]. 
Because of this importance, fisheries also represent some of the longest and best studied timeseries of population dynamics available to ecologists, as well as a correspondingly rich area for the development of theory and practice.
Methods developed in the study and management of fisheries are frequently applied to other areas of ecological management, from concepts such as maximum sustainable yield [], optimal control under observation uncertainty [@Memarzadeh2019], to adaptive management [@Walters], management strategy evaluation [@Punt], to the studies of conservation policy and international negotiations [@].

Mathematically, the ecological management may be understood as an optimal control problem - a manager must determine a sequence of actions (a "policy") to optimize some desired objective(s) and constraints (e.g. conservation goals, natural resource extraction, economic and social constraints.)
Given an appropriate representation of the underlying mechanistic processes involved, such sequential decision problems may be posed more precisely as Markov Decision Process [MDP, @Marescot2013; @Memarzadeh2019].
Such problems are notoriously difficult owing not only to the complexity of those underlying mechanisms alone, but to the exponentially branching consequences the decision-maker's actions have upon that system -- a game of chess is not won by thinking only one move ahead at a time. 
Thus, while MDP problems have been studied in a conservation context since they were first introduced more than six decades ago, their application has been highly constrained to simple scenarios.
Meanwhile, a class of machine learning techniques designed for MDP problems known as reinforcement learning (RL), has cracked open the door for exploring complex strategies in more realistic settings [@rl_intro].

- This is not a tutorial or a sales pitch for RL - which brings at least as many difficulties as advantages -- including being hard to use, easy to get stuck on far from optimal performance, which itself may be hard to diagnose, as we have already discussed in our tutorial. The method

has through the use of clever algorithms, deep neural networks, and massive amounts of data,

@@

```{r}

#| label: conceptual
#| echo: FALSE
#| message" FALSE
#| fig.align: 'center'
#| fig.cap: '(A) Classical management strategies often have the common aspect of reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $V1$). where the interaction between $V1$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$. An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive. For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2. 
#| (B) RL allows us to solve for quantitative decision strategies considering more complex ecosystem dynamics. For example, in this paper, we find harvest strategies for a three-species model using a model-free RL algorithm, PPO. 
#| (C) We show how the RL-based solution that considers all three species finds harvest policies with higher average reward and with (D) lower probability of harvest species collapse than the classical approaches, only considering the dynamics of V1.', 
#| out.width: '1\\linewidth', 
#| fig.pos: 'H'

#knitr::include_graphics("Figures/conceptual-figure.png")
```

@@

# Old Methods and Results

## Methods

Model-based controls, such as MSY, are known to have serious deficiencies when the model used to find the optimal control does not match reality.
As mentioned, for the particular case of fisheries this can lead to depletion of resources and affect severely the surrounding ecosystems.
Because of this, we have opted to compare the performance of several data-based approaches to fishery control (see Fig. \ref{fig:decision approaches}).

As shown in Fig. \ref{fig:decision approaches}, these data-based strategies rely on choosing a family of *approximating functions* for the policy, $\mathcal{P}:=\{\pi_\theta\}_\theta$, parametrized by some vector $\theta$.
These functions are deterministic policies, i.e. they map states to actions, or, specifically,
$$
  \pi_\theta : \mathbb{R}^n \to \mathbb{R}_+,
$$
where, as before, $n$ is the number of species.
The aim of the optimization process is to find a parameter choice $\theta^*$ for which $\pi_{\theta^*}$ is optimal within $\mathcal{P}$.

Our focus here will be comparing two data-based approaches: *constant mortality* and *DRL*.
Constant mortality is, as the name suggests, based on MSY.

For the constant mortality approach, we choose a linear harvest $\pi_F(N_t)= F X_t$, where the parameter $F$ is scalar, and where $N_t = (X_t,\ \dots\ )\in \mathbb{R}^n$.
This first control is intended to delineate how one might apply MSY to a real-world fishery.
We will also compare a TAC-inspired rule of thumb which uses, instead, the parameter value $F_{\text{TAC}} = 0.8 F^*$.
In App. \ref{app:tbd}, we discuss several other MSY-based approaches which yield similar results to the strategies just discussed.

In the DRL approach, we instead parametrize $\mathcal{P}$ using a neural network with two hidden layers of 64 neurons.
Our hypothesis is that the additional flexibility yielded by neural networks will allow the agent solutions that are better than MSY based approaches.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\linewidth]{figures/DecisionApproaches.pdf}
  \caption{
  \label{fig:decision approaches}
  TBD.
  }
\end{figure}

## Results

\textbf{Reword a bit.}
We trained a DRL agent to obtain a candidate policy
\begin{align*}
  \pi: (X_t,Y_t,Z_t) \mapsto h_t
\end{align*}
for eq. \eqref{eq:3d model}.
The agent could fully and precisely observe the state of the system at each time step and decide on its harvest based on this information. 
The agent was trained using the PPO algorithm, using a neural network with two hidden layers of 64 neurons to express $\pi$.
Details of the training process may be found in App. \ref{app:tbd}.
Our code is available at TBD.

Rewards were defined to be a compound of two contributions.
The first is economical: if at time $t$ a harvest of $h_t$ is secured, then the agents total reward is increased by $h_t$.
The second regards sustainability: if an extinction happens at time step $t$, then a penalty of $-50/t$ is incurred by the agent.
The choice of the constant $50$ in this rule was rather arbitrary: we set it to be a fourth of the maximal episode length ($200$ time steps), so as to encourage the agent to find policies that do not lead to extinctions below the time scale of around 100 time steps.
\textbf{Too long of a "sorry" message?}

We compare the performance of this agent to the performance of MSY and of an 80% TAC.

Our results suggest that DRL can find solutions that are more sustainable *and* more profitable than any constant mortality strategy.
Indeed, DRL finds solutions in which, for certain periods of time, no harvest is performed on the system.
These periods effectively allow the system to, in a sense, "reset" itself.
They lead to a distinctive oscillating behavior of the system state over time, with periods of relative depletion followed by periods of growth. 
Figure TBD exemplifies this point by comparing one times series of the system state obtained using each of these policies.

```{r timeseries_old, fig.cap="TBD.", out.width = "6.5in"}
policy_labeller <- function(variable,value){
  return(policy_labels[value])
}
msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(rep == 6) |>
    ggplot(aes(t, abundance, col = species)) + 
        geom_line()  + 
        scale_color_manual(values=c('palegreen3', 'coral1', 'deepskyblue4')) +
        facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
        labs(y = "abundance", x = "time") +
        theme(aspect.ratio = 2/3)
```

```{r rew_len_fig_old, fig.cap="TBD.", out.width = "6.5in"}
library(patchwork)

p1 <- msy_tac_ppo |> 
    group_by(rep, policy) |>
    filter(t==max(t), species == "X", rep < 50) |>
    ggplot(aes(rep, t)) + 
      geom_col() +
      facet_wrap(~factor(policy, levels = c("msy", "tac", "ppo")), labeller = policy_labeller ) +
      labs(y = "episode length", x = "repetition") +
      theme(axis.text.x = element_blank())

```{r state-space_old, fig.cap="TBD.", out.width = "6.5in"}
i_tac <- which.min(abs(actions - best_action * 0.8))

pmsy <- msy_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  filter(action == best_action) |>
  ggplot(aes(X, Z, color = harvest)) +
  geom_point() + sc +
  ggtitle("Opt. Const. Mortality")

ptac <- msy_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  filter(action == actions[[i_tac]]) |>
  ggplot(aes(X, Z, color = harvest)) +
  geom_point() + sc +
  ggtitle("80% Opt. Const. Mortality")

pppo <- ppo_df |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, harvest = action * X) |>
  ggplot(aes(X, Z, color = harvest)) +
  geom_point() + sc +
  ggtitle("DRL")

pesc <- opt_escapement |>
  mutate(X = X+1, Y = Y+1, Z = Z+1, 
        harvest = case_when(
          X - escapement > 0 ~ X - escapement,
          X - escapement <= 0 ~  0)
         ) |> 
  ggplot(aes(X, Z, color = harvest)) +
  geom_point() + sc +
  ggtitle("Opt. Const. Escapement")

```{r escapement_timeseries, fig.cap = "TBD", out.width = "6.5in"}

ppo_mod <- ppo_df |> 
  mutate(X = X+1, Y = Y+1, Z = Z+1, escapement = (1-action)*X, time = t)

avg_esc <- mean(ppo_mod$escapement)
# = 0.6201131

ppo_esc <- ppo_mod |>
  filter(rep == 6) |>
  group_by(rep) |>
  ggplot(aes(time, escapement), alpha = rep) +
  geom_line() +
  ylim(0,1) +
  geom_hline(aes(yintercept=avg_esc), color="red")+
  theme(aspect.ratio = 1/5)

ppo_esc
```

**Probably more for the discussion**
It is well known that following MSY policies can lead to over-fishing.
Our results reproduce this behavior and, furthermore, warn that using a TAC policy instead might not be enough to gain sustainability.
Figure \ref{rew_len_fig}-A summarizes episode length distribution for MSY, TAC and DRL.
While the DRL-derived policy does not lead to extinctions, both MSY and TAC do so, and thus their episode mean length is lower than the maximal episode length (in our case set at 200 years).
This difference is reflected on the profits earned by the agent, as shown in Fig. TBD-B.
